---
title: "class08: Breast Cancer Mini Project"
author: "Joseph Lo (PID: A18121493)"
format: pdf
toc: true
---

## Background

In today's class we will apply the methods and techniques clustering and PCA to help make sense of a real world breast cancer FNA biopsy data set.


## Data Import

We start by importing our data. It is a CSV file so we will use the `read.csv()` function.

```{r}
wisc.df <- read.csv("WisconsinCancer.csv", row.names=1)
```

have a wee peak at the first few rows:

```{r}
head(wisc.df)
```

Make sure to remove the first `diagnosis` column - I don't want to use this for my machine learning models. We will use it later to compare our results to the expert diagnosis.

```{r}
wisc.data <- wisc.df[,-1]
diagnosis <- wisc.df$diagnosis
```

## Exploratory data analysis

> Q1. How many observations are in this dataset?

569

```{r}
nrow(wisc.df)
```


> Q2. How many of the observations have a malignant diagnosis?

212

```{r}
sum(wisc.df$diagnosis == "M")
```

```{r}
table(wisc.df$diagnosis)
```

> Q3. How many variables/features in the data are suffixed with _mean?

10

```{r}
#colnames(wisc.data)

length(grep("_mean", colnames(wisc.data)))
```

## Principal Component Analysis

The main function here is `prcomp` and we want to make sure we set the optional argument `scale=TRUE`:

```{r}
wisc.pr <- prcomp(wisc.data, scale=TRUE)
summary(wisc.pr)
```

> Q4. From your results, what proportion of the original variance is captured by the first principal component (PC1)?

0.4427

> Q5. How many principal components (PCs) are required to describe at least 70% of the original variance in the data?

PC3

> Q6. How many principal components (PCs) are required to describe at least 90% of the original variance in the data?

PC7

Our main PCA "score plot" or "PC plot" of results:

```{r}
library(ggplot2)
```

```{r}
biplot(wisc.pr)
```

> Q7. What stands out to you about this plot? Is it easy or difficult to understand? Why

The black mass in the middle with red lines stands out to me the most. It is very difficult to understand because everything is packed together.


```{r}
ggplot(wisc.pr$x) +
  aes(PC1, PC2, col=diagnosis) +
  geom_point()
```

> Q8. Generate a similar plot for principal components 1 and 3. What do you notice about these plots?

The plots are distinguished between the blue and reds, and it is a organized scatter plot.


Scree-plot

```{r}
library(factoextra)
fviz_eig(wisc.pr, addlabels = TRUE)
```

```{r}
ggplot(wisc.pr$rotation) +
  aes(PC1,
      reorder(rownames(wisc.pr$rotation), PC1))+
  geom_col()
```

Collectively these two plot ("score plot" and "loadings plot") tell us that if cells nucleus are deeply indented ("concave"), irregular non circular ("compactness"), and have large "perimeter" values they tend to be malignant...

> Q9. For the first principal component, what is the component of the loading vector (i.e. wisc.pr$rotation[,1]) for the feature concave.points_mean? This tells us how much this original feature contributes to the first PC. Are there any features with larger contributions than this one?

```{r}
wisc.pr$rotation["concave.points_mean",1]
```

The component is the rowname and row. There are no features larger than the `concave.points_mean` because according to the bar graph, it is the largest one.

## Hierarchical clustering

First scale the data (with the `scale()` function), then calculate a distance matrix (with the `dist()` function). Then cluster with the `hclust()` function and plot:

```{r}
wisc.hclust <- hclust(dist(scale(wisc.data)))
```

> Q10. Using the plot() and abline() functions, what is the height at which the clustering model has 4 clusters?

height of 22

```{r}
plot(wisc.hclust)
abline(h=22, col="red")
```

You can also use `cutree()` function with a argument `k=4` rather than `h=height`

```{r}
wisc.hclust.clusters <- cutree(wisc.hclust, k=4)
table(wisc.hclust.clusters)
```


## Combining methods

Here we will take our PCA results and use those as input for clustering. In other words our `wisc.pr$x` scores that we plotted above (the main output from PCA - how the data lie on our new principal component axis/variables) and use a subset of these PCs that capture the most variance as input for `hclust()`

```{r}
pc.dist <- dist(wisc.pr$x[,1:3])
wisc.pr.hclust <- hclust(pc.dist, method = "ward.D2")
plot(wisc.pr.hclust)
```

Cut the dendrogram/tree into two main groups/clusters:

>Q12. Which method gives your favorite results for the same data.dist dataset? Explain your reasoning.

The results above is my favorite because, in my opinion, it looks the most organized and easy to dinstinguish.

```{r}
grps <- cutree(wisc.pr.hclust, k=2)
table(grps)
```

I want to know how the clustering in `grps` with values of 1 or 2 correspond the expert diagonosis

```{r}
table(grps, diagnosis)
```

My clustering **group 1** are mostly "M" diagnosis (179) and my clustering **group 2** are mostly "B" diagnosis (333)

24 FP
179 TP
333 TN
33 FN

> Q13. How well does the newly created hclust model with two clusters separate out the two “M” and “B” diagnoses?

It seperates them out pretty well because group 1 consists of mostly "M" and group 2 consists of mostly "B".

> Q14. How well do the hierarchical clustering models you created in the previous sections (i.e. without first doing PCA) do in terms of separating the diagnoses? Again, use the table() function to compare the output of each model (wisc.hclust.clusters and wisc.pr.hclust.clusters) with the vector containing the actual diagnoses.

The hierarchical clustering models also separates the diagnoses pretty well but it also includes a 2 and 4 section, which I don't know what it relates to.

Sensitivity TP/(TP+FN)

```{r}
179/(179+33)
```

Specificity TN/(TN+FP)
```{r}
333/(333+24)
```



## Prediction
```{r}
#url <- "new_samples.csv"
url <- "https://tinyurl.com/new-samples-CSV"
new <- read.csv(url)
npc <- predict(wisc.pr, newdata=new)
npc
```

```{r}
plot(wisc.pr$x[,1:2], col=grps)
points(npc[,1], npc[,2], col="blue", pch=16, cex=3)
text(npc[,1], npc[,2], c(1,2), col="white")
```

> Q16. Which of these new patients should we prioritize for follow up based on your results?

Patient 2 because they are more likely to be diagnosed as malignant than patient 1.











