---
title: "Class 12: RNASeq Galaxy Pt1/Population Analysis HW"
author: "Joseph Lo (PID: 18121493)"
format: pdf
toc: TRUE
---

# Section 1. Proportion og G/G in a population

Downloaded a CSV file from Ensemble <https://useast.ensembl.org/Homo_sapiens/Variation/Sample?db=core;r=17:39894781-39895412;v=rs8067378;vdb=variation;vf=959672880#373531_tablePanel>

Here we read this CSV file

```{r}
mxl <- read.csv("373531-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")

head(mxl)
```

```{r}
table(mxl$Genotype..forward.strand.)
```

```{r}
table(mxl$Genotype..forward.strand.)/nrow(mxl) *100
```

# Section 4 Population Analysis

How many samples do we have?

```{r}
expr <- read.table("rs8067378_ENSG00000172057.6.txt")

head(expr)
```

```{r}
nrow(expr)
```

```{r}
table(expr$geno)
```

```{r}
median(expr$exp[expr$geno=="A/A"])

median(expr$exp[expr$geno=="A/G"])

median(expr$exp[expr$geno=="G/G"])
```


> Q13: Read this file into R and determine the sample size for each genotype and their corresponding median expression levels for each of these genotypes.

The sample size for A/A is 108 with a median expression level of 31.2, A/G is 233 with a median expression level of 25.1, and G/G is 121 with a median expression level of 20.1. 

> Q14: Generate a boxplot with a box per genotype, what could you infer from the relative expression value between A/A and G/G displayed in this plot? Does the SNP effect the expression of ORMDL3?

According to the boxplot, The relative expression value for G/G is much lower than A/A, which makes me think that having a G/G is associated with having a reduced expression of the gene. The SNP does effect the expression of ORMDL3 with G/G being associated with a reduced expression of the gene.

```{r}
library(ggplot2)
```

Lets make a boxplot

```{r}
ggplot(expr) + aes(geno, exp, fill=geno) +
  geom_boxplot(notch=TRUE)
```






