---
title: "Class 18: Pertussis Mini-Project"
author: "Joseph Lo (PID: A18121493)"
format: pdf
---

# Background
Pertussis (whooping cough) is a common lung infection caused by the bacteria B. Pertussis. It can infect anyone but is most deadly for infants (under 1 year of age)

# CDC tracking data
The CDC tracks the number of Pertussis cases: 

```{r, echo=FALSE}
cdc <- data.frame(year = c(1922L,1923L,1924L,1925L,
                                          1926L,1927L,1928L,1929L,1930L,1931L,
                                          1932L,1933L,1934L,1935L,1936L,
                                          1937L,1938L,1939L,1940L,1941L,1942L,
                                          1943L,1944L,1945L,1946L,1947L,
                                          1948L,1949L,1950L,1951L,1952L,
                                          1953L,1954L,1955L,1956L,1957L,1958L,
                                          1959L,1960L,1961L,1962L,1963L,
                                          1964L,1965L,1966L,1967L,1968L,1969L,
                                          1970L,1971L,1972L,1973L,1974L,
                                          1975L,1976L,1977L,1978L,1979L,1980L,
                                          1981L,1982L,1983L,1984L,1985L,
                                          1986L,1987L,1988L,1989L,1990L,
                                          1991L,1992L,1993L,1994L,1995L,1996L,
                                          1997L,1998L,1999L,2000L,2001L,
                                          2002L,2003L,2004L,2005L,2006L,2007L,
                                          2008L,2009L,2010L,2011L,2012L,
                                          2013L,2014L,2015L,2016L,2017L,2018L,
                                          2019L,2020L,2021L,2022L,2023L, 2024L, 2025L),
                                cases = c(107473,164191,165418,152003,
                                          202210,181411,161799,197371,
                                          166914,172559,215343,179135,265269,
                                          180518,147237,214652,227319,103188,
                                          183866,222202,191383,191890,109873,
                                          133792,109860,156517,74715,69479,
                                          120718,68687,45030,37129,60886,
                                          62786,31732,28295,32148,40005,
                                          14809,11468,17749,17135,13005,6799,
                                          7717,9718,4810,3285,4249,3036,
                                          3287,1759,2402,1738,1010,2177,2063,
                                          1623,1730,1248,1895,2463,2276,
                                          3589,4195,2823,3450,4157,4570,
                                          2719,4083,6586,4617,5137,7796,6564,
                                          7405,7298,7867,7580,9771,11647,
                                          25827,25616,15632,10454,13278,
                                          16858,27550,18719,48277,28639,32971,
                                          20762,17972,18975,15609,18617,
                                          6124,2116,3044,7063,22538,21996)
       )
```

>Q. I want a plot of year vs cases

> Q1. With the help of the R “addin” package datapasta assign the CDC pertussis case number data to a data frame called cdc and use ggplot to make a plot of cases numbers over time.

```{r}
library(ggplot2)

ggplot(cdc) +
  aes(year, cases) +
  geom_point() +
  geom_line()
```

> Q. Add annotation lines for the major milestones of wP vaccination roll-out (1946) and the switch to the aP vaccine (1996).

>Q2. Using the ggplot geom_vline() function add lines to your previous plot for the 1946 introduction of the wP vaccine and the 1996 switch to aP vaccine (see example in the hint below). What do you notice?

After the introduction of the wP vaccine the rate of the infection decreased; however, after switching to the aP vaccine maybe it wasn't as strong but it is still pretty effective.

```{r}
ggplot(cdc) +
  aes(year, cases) +
  geom_point() +
  geom_line() +
  geom_vline(xintercept = 1946, col="blue", lty=2) +
  geom_vline(xintercept = 1996, col="red", lty=2) +
  geom_vline(xintercept = 2020, col="gray", lty=2)
```

>Q3. Describe what happened after the introduction of the aP vaccine? Do you have a possible explanation for the observed trend?

After the introduction of the aP vaccine there is a slight increase in the infection cases. Maybe the bacteria is evolving or that data gathering is evolving, making it easier to identify who is infected and including it into data sets with better accuracy. Also, access to vaccine may be tougher, causing less to get vaccinated.

>Q. Additional points for discussion: How are vaccines currently approved?

Vaccines are currently approved through companies like FDA so it will be harder to get modern vaccines improved. There is a more evolved process in approving vaccines and making sure there are no failures.

## Exploring CMI-PB data

The CMI-PB project <https://www.cmi-pb.org/>
The mission of CMI-PB is to provide the scientific community with a comprehensive, high-quality and freely accessible resource of Pertussis booster vaccination.

Basically, make available a large dataset on the immune response to Pertussis. They use a "booster" vaccination as a proxy for Pertussis infection. 

They make their data available as JSON format API. We can read this into R with the `read_json()` function from the **jsonlite** package:

```{r}
library(jsonlite)

subject <- read_json("https://www.cmi-pb.org/api/v5_1/subject", simplifyVector=TRUE)

head(subject)
```

> Q. How many aP and wP individuals are there in this `subject` table?


```{r}
table(subject$infancy_vac)
```

>Q4. How many aP and wP infancy vaccinated subjects are in the dataset?

87 aP and 85 wP

How many male/female
>Q5. How many Male and Female subjects/patients are in the dataset?

```{r}
table(subject$biological_sex)
```

112 Female 60 Male

>Q6. What is the breakdown of `race` and `biological sex` (e.g. number of Asian females, White males etc…)? Is this representative of the US population?

```{r}
table(subject$race, subject$biological_sex)
```

The breakdown is as shown above.

It is not representative of the US population because it is too little datasets.

We can read more tables from the CMI-PB database 

```{r}
specimen <- read_json("https://www.cmi-pb.org/api/v5_1/specimen", simplifyVector = TRUE)
ab_titer <- read_json("https://www.cmi-pb.org/api/v5_1/plasma_ab_titer", simplifyVector = TRUE)
```

```{r}
head(specimen)
```

```{r}
head(ab_titer)
```

To make sense of all this data we need to "join" (a.k.a. "merge" or "link") all these tables together. Only then will you know that a given Ab measurement (from the `ab_titer` table) was collected on a certain data (from the `specimen` table) from a certain wP or aP subject (from the `subject` table).

We can use **dplyr** and the `*_join()` family of functions to do this.

>Q9. Complete the code to join specimen and subject tables to make a new merged data frame containing all specimen records along with their associated subject details:

```{r}
library(dplyr)
meta <- inner_join(subject, specimen)
head(meta)
```

Let's do one more `innter_join()` to join the `ab_titer` with all our `meta` data.

>Q10. Now using the same procedure join meta with titer data so we can further analyze this data in terms of time of visit aP/wP, male/female etc.

```{r}
abdata <- inner_join(ab_titer, meta)
head(abdata)
```

> Q11. How many specimens (i.e. entries in abdata) do we have for each isotype?

```{r}
table(abdata$isotype)
```

The species amounts are shown above.

>Q. How many different "antigen" values are measured?


```{r}
table(abdata$antigen)
```

>Q12. What are the different $dataset values in abdata and what do you notice about the number of rows for the most “recent” dataset?

```{r}
table(abdata$dataset)
```

The most recent data set seems to have a jump in its number compared to previous years.

```{r}
igg <- abdata %>% filter(isotype == "IgG")
head(igg)
```


>Q13. Complete the following code to make a summary boxplot of Ab titer levels (MFI) for all antigens:

```{r}
library(ggplot2)
ggplot(igg) +
  aes(MFI_normalised, antigen) +
  geom_boxplot() +
  xlim(0,75) +
  facet_wrap(vars(visit), nrow=2)
```


Let's focus on IgG isotype

```{r}
igg <- abdata |>
  filter(isotype=="IgG")

head(igg)
```

Make a plot of `MFI_normalised` values for all `antigen` values.

```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen) +
  geom_boxplot()
```

>Q14. What antigens show differences in the level of IgG antibody titers recognizing them over time? Why these and not others?

The antigens "PT", "FTM2/3" and "FHA" appear to have the widest range of values. It is because the aP vaccines utilize these antibodies compared to wP which do not. This introduces more of these antibodies, thus the reason why we see more of them over time and different levels.

> Q. Is there a difference for these responses between aP and wP individuals?

```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen, col=infancy_vac) +
  geom_boxplot()
```

```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen) +
  geom_boxplot() +
  facet_wrap(~infancy_vac)
```

> Q.Is there a difference with time (i.e. before booster shot vs after booster shot)

```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen, col=infancy_vac) +
  geom_boxplot() +
  facet_wrap(~visit)
```

Lets finish this section by looking at the 2021 dataset IgG PT antigen levels time-course:


```{r}
## filter to 2021 dataset, IgG and PT only
ab.PT.21 <- abdata |> filter(dataset == "2021_dataset",isotype == "IgG",  antigen == "PT")

  ggplot(ab.PT.21) +
    aes(x=planned_day_relative_to_boost,
        y=MFI_normalised,
        col=infancy_vac,
        group=subject_id) +
    geom_point(alpha=0.4, size=1.5) +
    geom_line(alpha=0.25, linewidth=0.6) +
     stat_summary(fun=mean, geom="smooth", se=FALSE, linewidth=1.6, span=0.5, aes(group=infancy_vac, col=infancy_vac)) +
    geom_vline(xintercept=0, linetype="dashed") +
    geom_vline(xintercept=14, linetype="dashed") +
  labs(title="CMI-PB 2021 dataset IgG PT",
       subtitle = "Dashed lines indicate day 0 (pre-boost) and day 14 (post boost)") +
      theme_minimal(base_size=14)
```

>Q18. Does this trend look similar for the 2020 dataset?

Yes this trend looks similar to the 2020 dataset.

```{r}
ab.PT.20 <- abdata |> filter(dataset == "2020_dataset",isotype == "IgG",  antigen == "PT")


  ggplot(ab.PT.21) +
    aes(x=planned_day_relative_to_boost,
        y=MFI_normalised,
        col=infancy_vac,
        group=subject_id) +
    geom_point() +
    geom_line() +
    geom_vline(xintercept=0, linetype="dashed") +
    geom_vline(xintercept=14, linetype="dashed") +
  labs(title="2020 dataset IgG PT",
       subtitle = "Dashed lines indicate day 0 (pre-boost) and 14 (apparent peak levels)")
```

